---
title: "Qualitative Activity Recognition of Weight Lifting Exercises"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## What You Should Submit

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## How the Model Was Built

Participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. [1]

In this model, the outcome variable is "classe", a factor variable with 5 levels. Once the dataset is clean and organised, predictions based on decision trees and random forests will be performed and assessed based on their accuracy. 

## Cross Validation and the Out of Sample Error

The training data set will be split into 70% subTrain ans 30% subTest. The models will be built using the subTrain data set and run on the subTest data set in order to estimate how these models would perform on the independent Testing data set. 

The out of sample error rate is the error you get on a new data set. This can be estimated by the error rate we get on the subTest data set, as 1 - accuracy.

## Model Justification

Decision trees and random forests algorithms have the ability to detect the valuable features for classification. In addition, the data set is large enough for us to split into subTrain and subTest and perform cross-validation. 

## Data Preparartion

Install the following packages before running below code: caret, randomForest and rpart.

```{r, echo=TRUE}
library(caret)
library(randomForest)
library(rpart)
set.seed(1333)

# Getting and Cleaning Data
trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))

training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

# Remove unnecesary variables
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]

dim(training)
dim(testing)
```

## Creating subTrain and subTest partitions for cross validation

The training data set will be split into subTrain (70%) and subTest (30%) using the createDataPartition function in the caret package.

```{r, echo=TRUE}
inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
subTrain <- training[inTrain, ]
subTest <- training[-inTrain, ]
dim(subTrain)
dim(subTest)
```

## Prediction 1: Decision Tree

```{r, echo=TRUE}
DT <- rpart(classe ~ ., data=subTrain, method="class")
DTpred <- predict(DT, subTest, type = "class")
confusionMatrix(DTpred, subTest$classe)
```

## Prediction 2: Random Forests

```{r, echo=TRUE}
RF <- randomForest(classe ~. , data=subTrain, method="class")
RFpred <- predict(RF, subTest, type = "class")
confusionMatrix(RFpred, subTest$classe)
```

## Conclusion

With a far higher accuracy (99%) for the Random Forests model on the subTest data set than the Decision Tree model (75%), I have selected the Decision Tree (DT), as my prediction model.

## Submission

```{r, echo=TRUE}
predictions <- predict(DT, testing, type="class")
predictions

submission = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("test_case",i,".txt")
    write.table(x[i], file=filename,quote=FALSE,row.names=FALSE,col.names = FALSE)
  }
}

submission(predictions)
```

## References

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz4jzDXY2Ca